---
title: "Statistical Analysis OES Paper"
author: "Amy Hatton"
date: "29/06/2021"
output: pdf_document
fontsize: 12pt
mainfont: Calibri
---
# Ostrich eggshell beads from Ga-Mohana Hill North Rockshelter, southern Kalahari, and the implications for understanding social networks during Marine Isotope Stage 2

Authors: Amy Hatton, Benjamin Collins, Benjamin J. Schoville, Jayne Wilkins

## Code for statistical analysis in this paper
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#libraries
library(dplyr)
library(janitor)
```

### Chi-squared test to compare levels of burning between OES frags and beads

Read in OES data and process
```{r}
oesbead_data <- read.csv("../data/ghn_oesbeads.csv")
oes_data <- read.csv("../data/ghn_oes.csv")
```
Filter out the DBGS points
```{r}
# filter out only the DBGS finds
oes_beads_dbgs <- oesbead_data %>% 
  filter(Strat.Agg == "DBGS") %>% 
  clean_names() %>% 
  mutate(colour = color) 
oes_dbgs <- oes_data %>% 
  filter(Strat.Agg == "DBGS")%>% 
  clean_names()
```

Get the range and sd of some of the technological variables
```{r}
range(oes_dbgs$length_mm)
sd(oes_dbgs$length_mm)

range(oes_dbgs$width_mm)
sd(oes_dbgs$width_mm)

range(oes_dbgs$weight_g)
sd(oes_dbgs$weight_g)
diam_1 <- oes_beads_dbgs %>% 
  filter(bead_or_preform == "bead") 
sd(diam_1$max_ext_di)
sd(diam_1$max_aper_di)
sd(oes_beads_dbgs$max_thick)
```

Combine OES frags and OES beads as one df
```{r}
oes_beads_dbgs <- oes_beads_dbgs %>% 
  mutate(type = "bead_or_preform") %>% 
  dplyr::select(lot, find, type, colour)
  
oes_dbgs <- oes_dbgs %>% 
  mutate(type = "fragment")%>% 
  dplyr::select(lot, find, type, colour)

#combine the two into 1 df
dbgs_oes <- rbind(oes_beads_dbgs,oes_dbgs)
```

Check that the levels for colour are clean
```{r}
dbgs_oes$colour <- as.factor(dbgs_oes$colour)
levels(dbgs_oes$colour)

#need to subsume brown into unburnt (brown is just staining from sediment)
levels(dbgs_oes$colour) <- c("black"  , "unburnt"  , "red"   ,  "unburnt" ,"yellow" )
```
### Chi-squared test

Create contigency table
```{r}
library(MASS)       # load the MASS package 
tbl <- table(dbgs_oes$colour, dbgs_oes$type)
```
Run chi-sqaured test
```{r}
chisq.test(tbl)
```
### Anova for oes bead mean diameter

Read in the data
```{r}
sa_beads <- read.csv("../data/sa_oes_bead_table.csv")

sa_beads <- clean_names(sa_beads)
```
Process the data
```{r}
#group by site
bead_diam <- sa_beads %>% 
  dplyr::select(site, site_abb,mean_diameter_mm) %>% 
  filter(!is.na(mean_diameter_mm)) %>% 
  group_by(site) %>% 
  mutate(diameter = mean(mean_diameter_mm)) 

bead_diam_mean <- sa_beads %>% 
  dplyr::select(site, site_abb,mean_diameter_mm) %>% 
  filter(!is.na(mean_diameter_mm)) %>% 
  group_by(site) %>% 
  mutate(diameter = mean(mean_diameter_mm)) %>% 
  dplyr::select(!mean_diameter_mm) %>% 
  distinct()

```
Conduct anova test
```{r}
# Compute the analysis of variance
res.aov <- aov(mean_diameter_mm ~ site, data = bead_diam)
# Summary of the analysis
summary(res.aov)
#plot(res.aov)

#Tukey pairwise comparisons
TukeyHSD(res.aov)
```
Check which sites have more than 50 beads
```{r}
ba <- sa_beads %>% 
  mutate(bead_total = beads_finished + beads_preforms ) %>% 
  filter(bead_total > 50)

range(ba$bead_total)
```
### Spearman rank order correlation

Run Spearman rank order correlation on ostrich prevalence and bead size
```{r}
library(raster)
library(sf)
# read in oes distribution raster
oes_dist <- raster("../data/ostrich_distribution.tif")

sa_beads_sf <- st_as_sf(sa_beads, coords = c("long","lat" ),
                     crs = 4326)


bead_diam_sf <-  sa_beads_sf %>% 
  dplyr::select(site, site_abb,mean_diameter_mm) %>% 
  filter(!is.na(mean_diameter_mm)) %>% 
  group_by(site) %>% 
  mutate(diameter = mean(mean_diameter_mm)) %>% 
  dplyr::select(!mean_diameter_mm) %>% 
  distinct(.keep_all=TRUE)

#get the raster values (ostrich prevalence) for a 5km area around each site

ras_value <- raster::extract(oes_dist, bead_diam_sf, buffer= 5000, fun=mean)

ras_value1 <- raster::extract(oes_dist, sa_beads_sf, buffer= 5000, fun=mean)

val_points <- cbind(bead_diam_sf, ras_value)
```




